---
title: "Getting Started with landconn"
author: "Bill Peterman"
output:
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with landconn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4.25
)
```

# Introduction

This vignette walks through the two things `landconn` is built to do: measure
connectivity among discrete habitat patches with an incidence function, and measure
connectivity across a continuous resistance surface with Circuitscape. It is written for
readers who are new to the package. You do not need prior connectivity modeling
experience, but a basic familiarity with R and with spatial data helps.

The incidence function workflow runs anywhere R does. The Circuitscape workflow needs a
local Julia installation, so those sections are shown but not run here.

```{r libraries}
library(landconn)
```

# Patch connectivity with the incidence function

The incidence function approach treats the landscape as a set of point locations (patches
or sample sites) and asks, for each one, how well connected it is to all the others. Two
ingredients drive the calculation: the pairwise distances between sites, and `alpha`, a
parameter describing how far the organism typically disperses.

## Build a distance matrix

We start with some site coordinates. In practice these are your real sampling locations;
here we simulate ten of them.

```{r coords}
set.seed(123)
loc <- data.frame(x = runif(10, 0, 10),
                  y = runif(10, 0, 10))

d_mat <- dist_mat(loc)
```

`dist_mat()` returns a `land_dist` object, which is a square matrix of straight-line
distances that also knows how to summarize and plot itself.

```{r dist-summary}
summary(d_mat)
```

**How to read the `summary()` output:** the values describe the spread of the unique
pairwise distances between sites (the lower triangle of the matrix, so each pair is
counted once). The minimum is the closest pair of sites and the maximum is the farthest.
These distances are in the same units as your coordinates, so they set the scale on which
`alpha` should be interpreted in the next step.

```{r dist-plot, fig.alt = "Heatmap of pairwise distances between ten sites."}
plot(d_mat)
```

**How to read this plot:** each cell is the distance between a pair of sites, with darker
or lighter shading for shorter or longer distances depending on your color scale. The
diagonal is zero (each site is zero distance from itself). Clusters of short distances
point to groups of sites that sit close together on the landscape.

## Choosing `alpha`

`alpha` is the average dispersal distance of your organism, in the same units as the
distances in `d_mat`. It enters a negative exponential that converts each pairwise
distance into a probability of connection:

$$ p_{ij} = \exp\!\left(-\frac{1}{\alpha}\, d_{ij}\right) $$

A small `alpha` means connection probability falls off quickly with distance, so only
near neighbors contribute. A large `alpha` means even distant sites stay connected. Set
`alpha` from what you know about the species' movement, and remember that it is on the
same distance scale as your coordinates.

```{r alpha}
alpha <- 3
```

## Calculate connectivity

`ifc()` offers four models. All share the distance-decay core above; they differ in
whether and how they weight by patch area.

```{r models}
## Model 1: distance only
c1 <- ifc(alpha, d_mat, model = 1)

## Model 1, scaled to a maximum of 1
c1s <- ifc(alpha, d_mat, model = 1, scale = TRUE)
```

The four models encode different ecological hypotheses:

| Model | Weighting | Hypothesis |
|-------|-----------|------------|
| 1 | Distance only | Connectivity declines as a site becomes more isolated |
| 2 | Contributing patch area | Larger neighboring patches send more individuals |
| 3 | Focal patch area | Larger focal patches attract more individuals |
| 4 | Both | Larger patches both send and attract more individuals |

Models 2 to 4 need a patch area for each site.

```{r area-models}
p_a <- runif(10, 5, 50)

c2 <- ifc(alpha, d_mat, model = 2, patch_area = p_a)
c3 <- ifc(alpha, d_mat, model = 3, patch_area = p_a)
c4 <- ifc(alpha, d_mat, model = 4, patch_area = p_a)
```

You can restrict which sites contribute by passing `occ_sites`, a 0/1 vector marking
occupied sites. Only occupied sites then contribute to the connectivity of others.

```{r occupancy}
occ_sites <- c(0, 0, 1, 1, 0, 1, 0, 0, 1, 1)
c4b <- ifc(alpha, d_mat, model = 4, occ_sites = occ_sites, patch_area = p_a)
```

## Reading the connectivity output

The object that comes back is an `ifc` object. Printing it shows the model context and
the per-site values.

```{r print-ifc}
c4
```

```{r summary-ifc}
summary(c4)
```

**How to read the `summary()` output:** the header restates the model, `alpha`, the area
transform, and whether the values were scaled. The distribution block shows the spread of
relative connectivity across sites. The most and least connected sites are listed by name
(here, their row numbers), which is usually what you act on: the least connected sites are
candidates for restoration or closer monitoring.

Connectivity values are *relative*. They are meaningful in comparison with one another
within a single call, not as absolute numbers. A site with twice the value of another is
roughly twice as connected under that model. Because the models are on different scales
(Model 4 multiplies by area twice, for example), do not compare raw values across models.
To compare sites on a common 0 to 1 scale, use `scale = TRUE`, or compare the *rankings*
rather than the magnitudes.

```{r plot-ifc, fig.alt = "Barplot of relative connectivity by site, sorted from most to least connected."}
plot(c4)
```

**How to read this plot:** each bar is one site's relative connectivity, sorted from most
to least connected. The tallest bars are well-connected hubs; the shortest are the most
isolated sites in the network.

## Transforming patch area

By default (`area_transform = "none"`) Models 2 to 4 use patch area directly, matching the
standard incidence function equations. Setting `area_transform = "log"` substitutes the
log of area, which dampens the pull of very large patches. Use `"log"` only when your
areas are comfortably greater than 1, because the log of an area below 1 is negative.

```{r transform}
c2_none <- ifc(alpha, d_mat, model = 2, patch_area = p_a)
c2_log  <- ifc(alpha, d_mat, model = 2, patch_area = p_a, area_transform = "log")

## The two weightings rank sites differently
cor(c2_none, c2_log, method = "spearman")
```

**How to read this:** a Spearman correlation well below 1 tells you the two area
transforms reorder the sites, so the choice of transform is a modeling decision worth
making deliberately rather than by default.

## Estimating the dispersal scale (alpha)

So far we set `alpha` by hand. Often you would rather let data tell you the scale. If you
have a response measured at each site (occupancy, abundance, colonization), `ifc_optim()`
finds the `alpha` whose connectivity best predicts that response in a generalized linear
model. This is a *scale of effect* optimization: for each candidate `alpha` it recomputes
connectivity, fits the model, and keeps the `alpha` that maximizes the fit.

To see it work, we simulate site abundances that depend on connectivity at a known
`alpha` of 4, then try to recover it. The predictor named `connectivity` in the formula
is reserved: `ifc_optim()` generates it internally at each `alpha`.

```{r optim-sim}
set.seed(1)
opt_loc <- data.frame(x = runif(120, 0, 20), y = runif(120, 0, 20))
opt_d   <- dist_mat(opt_loc)

## True connectivity at alpha = 4, plus a continuous response with modest noise
true_conn <- as.numeric(ifc(4, opt_d, model = 1, scale = TRUE))
opt_dat   <- data.frame(abundance = 2 + 8 * true_conn + rnorm(120, 0, 0.5))

fit <- ifc_optim(abundance ~ connectivity, data = opt_dat,
                 dist_mat = opt_d, model = 1, family = gaussian)
fit
```

**How to read the output:** `alpha` is the estimated dispersal scale, in the distance
units of your coordinates. It is the distance over which connection probability falls to
about 37% (that is, `exp(-1)`) of its maximum. The profile confidence interval comes from
the curvature of the log-likelihood around the estimate.

```{r optim-plot, fig.alt = "Log-likelihood profile against alpha with the estimate and confidence interval marked."}
plot(fit)
```

**How to read this plot:** the curve is the log-likelihood across candidate `alpha`
values (note the log x axis). The solid red line is the estimate and the dotted red lines
the confidence interval. A sharp peak, as here, means the data identify `alpha` well. A
flat profile would warn you that the data carry little information about the scale, so the
estimate should not be trusted. Binary occupancy responses, in particular, often give
flat profiles; a continuous or count response is far more informative.

For a confidence interval that does not rely on the shape of the profile, add a parametric
bootstrap. It simulates new responses from the fitted model, holding the site network
fixed, and re-optimizes `alpha` for each.

```{r optim-boot, eval = FALSE}
fit_b <- ifc_optim(abundance ~ connectivity, data = opt_dat,
                   dist_mat = opt_d, model = 1, family = gaussian,
                   n_boot = 200)
fit_b$alpha_ci_boot
```

One caution: the model treats connectivity as a fixed predictor, so the standard errors
that `summary(fit)` reports for the `connectivity` coefficient are conditional on the
estimated `alpha` and are slightly optimistic. The interval on `alpha` itself comes from
the profile or the bootstrap, not from those standard errors.

## Comparing incidence models

You usually do not know which of the four `ifc()` models, or which area transform, best
describes your system. Pass a vector of models (and, if you like, both area transforms)
and `ifc_optim()` fits the best `alpha` for each combination and ranks them in a model
selection table.

```{r modsel}
opt_pa <- runif(120, 5, 50)
fits <- ifc_optim(abundance ~ connectivity, data = opt_dat, dist_mat = opt_d,
                  model = 1:4, patch_area = opt_pa,
                  area_transform = c("none", "log"), family = gaussian)
fits
```

**How to read the table:** each row is one model with its own optimized `alpha`. `K` is
the number of parameters, counted correctly: the GLM coefficients, the dispersion
parameter (gaussian estimates one), and the optimized `alpha`. Ranking is by AICc, with
`delta` the distance from the best model and `weight` the Akaike weight (the models'
weights sum to 1 and express relative support). Models within about 2 `delta` units of
the best have comparable support. Because every model here estimates `alpha`, the AICc
already accounts for that flexibility, so the comparison is fair.

```{r modsel-plot, fig.alt = "Log-likelihood profiles for each ifc model overlaid on one axis."}
plot(fits)
```

**How to read this plot:** each curve is one model's `alpha` profile. The model whose peak
sits highest, after the AICc penalty for parameters, is the best supported. Drill into it
with `fits$best`, which is an ordinary fit with its own `summary()` and `plot()`.

The information criteria are also available directly. Because the `K` for an `ifc_optim`
fit counts `alpha`, you can compare a connectivity model against a null model with no
connectivity term in one call:

```{r ic, eval = FALSE}
AICc(fits$best)                                   # AICc of the best model
null <- glm(abundance ~ 1, data = opt_dat)        # no connectivity
AIC(fits$best, null)                              # alpha is counted for the ifc fit
```

## Other model classes

The formula interface above fits a GLM. To use another model class, fit it once with a
placeholder `connectivity` term and pass the fitted object; `ifc_optim()` refits it across
`alpha`. This works for negative binomial models (`MASS::glm.nb`) and for `unmarked`
occupancy and abundance models, where `connectivity` is a site covariate. The parameter
counts in the information criteria include each model's own parameters (the negative
binomial dispersion, the detection parameters of an occupancy model) plus `alpha`.

```{r engines, eval = FALSE}
## Negative binomial counts
library(MASS)
nb_dat <- data.frame(count = counts, connectivity = 0)   # placeholder term
nb0 <- glm.nb(count ~ connectivity, data = nb_dat)
ifc_optim(nb0, dist_mat = d, model = 1)

## unmarked occupancy: connectivity goes in the siteCovs
library(unmarked)
siteCovs(umf)$connectivity <- 0                          # placeholder
occ0 <- occu(~1 ~ connectivity, data = umf)
ifc_optim(occ0, dist_mat = d, model = 1)
```

**How to read this:** the result is the same `ifc_optim` object as before, with `print()`,
`summary()`, `plot()`, and the information criteria, so everything in this vignette applies
unchanged. Keep the template simple: complex offsets or weights may not survive the refit.

# Resistance-surface connectivity with Circuitscape

The second approach models the whole landscape as a resistance surface, where every cell
carries a cost to movement, and uses Circuitscape to compute current flow across it.
`landconn` drives Circuitscape from R through `JuliaConnectoR`.

The package ships a small example resistance raster.

```{r raster}
f <- system.file("extdata/resist.tif", package = "landconn")
r <- terra::rast(f)
terra::plot(r)
```

**How to read this map:** cell values are resistance to movement, where higher values are
harder to cross. Circuitscape treats low-resistance cells as easier paths, so current
concentrates where movement is cheapest.

## Running Circuitscape

These calls require Julia and the Circuitscape library, so they are shown but not run. The
first time, install the Julia libraries once with `julia_packages()`.

```{r jl-setup, eval = FALSE}
## Point to the Julia 'bin' folder on your machine; install libraries the first time only
julia_packages(julia_path = "C:/Programs/Julia-1.10.5/bin", install_libraries = TRUE)
```

`run_cs()` runs a Circuitscape analysis between a set of locations and returns the
cumulative current map as a `SpatRaster`.

```{r run-cs, eval = FALSE}
jl_home <- "C:/Programs/Julia-1.10.5/bin"

pts <- terra::vect(rbind(c(250, 250),
                         c(250, 1750),
                         c(1750, 1750),
                         c(1750, 250)))

cs_out <- run_cs(JULIA_HOME = jl_home,
                 r = r,
                 input_locs = pts,
                 output_dir = tempdir(),
                 output_name = "demo",
                 is_resistance = TRUE,
                 silent = TRUE)
terra::plot(cs_out)
```

**How to read the result:** the returned raster is a cumulative current map. Bright,
high-current cells are pinch points where movement is funneled and where connectivity is
most vulnerable to loss. Diffuse current indicates many redundant pathways.

`omni_cs()` wraps the same machinery to produce an omnidirectional current map, following
Koen et al. (2014). It surrounds the landscape with nodes and runs an all-to-one analysis,
which highlights connectivity without assuming particular source and destination points.

```{r omni-cs, eval = FALSE}
omni_r <- omni_cs(JULIA_HOME = jl_home,
                  r = r,
                  n_nodes = 20,
                  is_resistance = TRUE,
                  silent = TRUE)
terra::plot(omni_r)
```

# Quick reference

The full incidence function workflow in one block:

```{r quickref, eval = FALSE}
library(landconn)

## 1. Site coordinates -> distance matrix
loc   <- data.frame(x = runif(10, 0, 10), y = runif(10, 0, 10))
d_mat <- dist_mat(loc)                 # a `land_dist` object

## 2. Set dispersal scale and (optionally) patch areas and occupancy
alpha     <- 3                         # average dispersal distance, in coordinate units
p_a       <- runif(10, 5, 50)          # patch areas
occ_sites <- c(0,0,1,1,0,1,0,0,1,1)    # 0/1 occupancy

## 3. Connectivity under the chosen model
conn <- ifc(alpha, d_mat, model = 4,   # both contributing and focal area
            patch_area = p_a,
            occ_sites = occ_sites)

## 4. Read the result
conn                                   # per-site relative connectivity
summary(conn)                          # distribution + most/least connected sites
plot(conn)                             # sorted barplot
```

## Summary of key functions

| Function | Purpose |
|----------|---------|
| `dist_mat()` | Build a pairwise distance matrix (`land_dist`) from coordinates |
| `lower()` | Extract the lower triangle of a distance matrix as a vector |
| `ifc()` | Incidence function connectivity, four models, returns an `ifc` object |
| `ifc_optim()` | Estimate the incidence scale `alpha` by optimizing GLM fit; compare models |
| `ifc_modsel()` | Rank `ifc_optim` fits by AIC, AICc, or BIC, counting `alpha` |
| `run_cs()` | Run Circuitscape between locations across a resistance surface |
| `omni_cs()` | Omnidirectional current map (Koen et al. 2014) |
| `julia_packages()` | Start Julia and load the Circuitscape libraries |

# References

Koen, E. L., Bowman, J., Sadowski, C., and Walpole, A. A. (2014). Landscape connectivity
for wildlife: development and validation of multispecies linkage maps. *Methods in
Ecology and Evolution*, 5(7), 626-633. <doi:10.1111/2041-210X.12197>
